---
title: 'Basic Skills in R Markdown: the pdf file'
output:
  pdf_document: default
  html_document: default
  word_document: default
subtitle: Ziv Shkedy (October, 2021)
layout: page
toc: yes
---


---
<!-- output: html_document -->
<!--   use_bookdown: TRUE -->  
--- 


```{r,echo=FALSE}
library(e1071)
library(lattice)
library(ggplot2)
library(mvtnorm)
```

\newpage


# Introduction

This document provides basic tools to produce a html file using R markdown. The best way to use this document is to run the file in R studio and then read the <tt>.Rmd</tt> file to see how the output was created. The file can be used to produce a very basic html document an you can add later more components to you document.

# Sections and subsections

This is a an example of a R markdown file that produces htnl output.  This is a section in the document.

## Subsection

This text apears in a subsection

### Subsubsection 

This text is a part of a subsection.



# Including R code

## Print R code and output

This is an example how to include R code and output in the document. We use the <tt>airquality</tt> data as an example.


```{r}
x<-na.omit(airquality$Ozone)
print(x) 
```

## Do not print the R code but print the output

If we do not want to print the R code, but we want to see the output use the option <tt>echo=FALSE</tt>:

```{r,echo=FALSE}
x<-na.omit(airquality$Ozone)
print(x) 
```


# Items

This is a text that contains items:

 * Item 1
 * Item 2
 * Item 3 
 * Item 4
 
 
# Use R as a part of your text

## Example: summary statistics

The mean Ozone level is 

```{r}
x<-na.omit(airquality$Ozone)
mean(x) 
```

with variance is equal to

```{r}
var(x) 
```

## Graphical displays in the document

A histogram for the Ozone level can be produced using the function <tt>qplot</tt> with the option <tt>geom = "histogram"</tt>:

```{r}
Ozone.R<-data.frame(x)
qplot(x, data = Ozone.R, geom = "histogram", binwidth = 0.1)
```


To add a caption to the figure we use <tt>{r figchp1,fig.cap="Sepal length (III)"}</tt>.

```{r figchp1,fig.cap="Sepal length (III)"}
Ozone.R<-data.frame(x)
qplot(x, data = Ozone.R, geom = "histogram", binwidth = 0.1)
```

We can refer to the figure from the text in the document. For example, Figure \@ref(fig:figchp2)  presents a histogram that was produced using the function <tt>qplot()</tt> function.

```{r figchp2,fig.cap="Sepal length (III)"}
Ozone.R<-data.frame(x)
qplot(x, data = Ozone.R, geom = "histogram", binwidth = 0.1)
```

# How to add a link to your document

Materials about R markdown are widely available online. 

 * For a YouTube tutorial about R markdown by 
   Jalayer Academy see [YTRmd1](https://www.youtube.com/watch?v=tKUufzpoHDE).
 * For a second YouTube tutorial about R markdown  by 
   Roger Peng see [YTRmd2](https://www.youtube.com/watch?v=DNS7i2m4sB0).
 
A usefull link to a R Markdown:

 * For a free online book about R markdown see see [RMDbook](https://bookdown.org/yihui/rmarkdown/).

# How to create a math formula

To create a math formula, for example a linear regression model of the form

$$y_{i}=\alpha+\beta \times x_{i} + \varepsilon_{i},$$
we need to use $LaTeX$ syntax. 


# Just do it

## Analysis of the cars data

Fit a linear regression model to the <tt>cars</tt> datasets in which the response is the stopping distance and the predictor is the car speed.

## Expected outout

Write a short report with the following structure

  * Introduction
  * The cars data (including a scapter plot of the data)
  * Modeling (formulate the model for the cars data)
  * Application to the data: present the results including the R object with the regression output. Plot os the data   
    and fitted model and isgnostic plots.
    
  
In your report, include the R code as a part of the text. Produce both html and pdf outputs.



