---
title: 'Case study 1: analysis of the daily average wind speed in New York in 1973'
output:
  pdf_document: default
  html_document: default
  word_document: default
subtitle: Foundations for inference using R
layout: page
---


---
<!-- output: html_document -->
<!--   use_bookdown: TRUE -->
--- 


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,	
                      warning = FALSE,
                      eval = TRUE,
                      tidy = FALSE)
library(knitr)
library(tidyverse)
library(deSolve)
library(minpack.lm)
library(ggpubr)
library(readxl)
library(gamlss)
library(data.table)
library(grid)
library(png)
library(nlme)
library(gridExtra)
library(mvtnorm)
library(e1071)
library(lattice)
library(ggplot2)
library(dslabs)
library(NHANES)
library(dplyr)
library(nasaweather)
library(ggplot2)
library(gganimate)
library(av)
library(gifski)
library(foreach)
library("DAAG")
```


# Case study 1: The wind speed in the airquality dataset

## Exploratory analysis of the daily average of the wind speed

The <tt>airquality</tt> dataset is a R object gives information about 153 daily air quality measurements ($n=153$) in New York, May to September 1973. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
dim(airquality)
head(airquality)
```

The variable of primary interest, <tt>Wind</tt>, is the average wind speed in miles per hour at 0700 and 1000 hours at LaGuardia Airport. We use the R package <tt>ggplot2</tt> to explore the data. Figure 1 shows histogram of wind speed.

```{r, echo=TRUE, message=FALSE, warning=FALSE,fig.cap="Figure 1: Histogram of wind speed."}
ggplot(airquality, aes(x = Wind)) +
  geom_histogram(fill = "skyblue", color = "black")+
  ylab("Frequency")
```

The sample mean ($\bar{x}$) and standard deviation ($s$) of the wind speed are 9.957516 and 3.523001, respectivly.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
wind<-airquality$Wind
M.wind<-mean(wind)
SD.wind<-sqrt(var(wind))
M.wind
SD.wind
```

## A $95\%$ confidence interval for the population mean

We constrct a $95\%$ confidence interval for the population mean using the R fucntion <tt>z.test</tt>. We assume that $\sigma=s$ and the function uses the $N(0,1)$ to select the critical value $Z_{\alpha}$. For a $95\%$ confidence interval, $Z_{\alpha}=1.96$.
A $95\%$ confidence interval for the wind speed is [9.399284; 10.515749]. This means that we are $95\%$ confident that the true average wind speed lies within this range.


## Test of hypothesis about the population mean

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(TeachingDemos)
wind=na.omit(airquality$Wind)
z.test(wind,sd=SD.wind)
```

Testing the hypotheses whether the wind speed is equal to 9 versus a two-sided alternative hypothesis at the significant level of 0.05 can be formulated by:

$$H_{0}:\mu=9\;\;\mbox{Vs.}\;\; H_{1}:\mu \ne 9.$$

We use the <tt>z.test()</tt> function and specify <tt>mu=9$.

```{r, echo=TRUE, message=FALSE, warning=FALSE}

z.test(wind, SD.wind, mu=9)
```

Since p-value = 0.0007742 which is much smaller than $\alpha = 0.05$, there is sufficient evidence to say that the mean of the wind speed is not equal to 9.

