---
title: "TMLE"
output: html_document
date: "2023-11-27"
---
## Page: https://www.khstats.com/blog/tmle/tutorial
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages("gt")
# Q
install.packages("earth")
# install.packages("ranger")
library(tidyverse) # for data manipulation
library(gt) # for table printing
library(SuperLearner) # for ensemble learning
library(earth)
library(ranger)
set.seed(7) # for reproducible results
```
## Generate the Data
## W1, W2 
```{r}
generate_data <- function(n){ 
    W1 <- rbinom(n, size=1, prob=0.2) # binary confounder
    W2 <- rbinom(n, size=1, prob=0.5) # binary confounder
    W3 <- round(runif(n, min=2, max=7)) # continuous confounder
    W4 <- round(runif(n, min=0, max=4)) # continuous confounder
    A  <- rbinom(n, size=1, prob= plogis(-0.2 + 0.2*W2 + log(0.1*W3) + 0.3*W4 + 0.2*W1*W4)) # binary treatment depends on confounders
    Y <- rbinom(n, size=1, prob= plogis(-1 + A - 0.1*W1 + 0.2*W2 + 0.3*W3 - 0.1*W4 + sin(0.1*W2*W4))) # binary outcome depends on confounders
    return(tibble(Y, W1, W2, W3, W4, A))
}

n <- 1000
dat_obs <- generate_data(n) # generate a data set with n observations

dat_obs %>%
  head() %>%
  gt() %>%
  tab_header("Simulated data set.")
```
```{r}
sl_libs <- c('SL.glmnet', 'SL.ranger', 'SL.earth', 'SL.glm') # a library of machine learning algorithms (penalized regression, random forests, and multivariate adaptive regression splines)
```

Fit a super-learner:  Is a linear combination of Learners, with estimated weights.
```{r}
Y <- dat_obs$Y
W_A <- dat_obs %>% select(-Y) # remove the outcome to make a matrix of predictors (A, W1, W2, W3, W4) for SuperLearner
Q <- SuperLearner(Y = Y, # Y is the outcome vector
        X = W_A, # W_A is the matrix of W1, W2, W3, W4, and A
        family=binomial(), # specify we have a binary outcome
        SL.library = sl_libs) # specify our superlearner library of LASSO, RF (ranger), and MARS (earth)

```
Form the predictions for 3 cases( A, A=0, A=1) 
```{r}
Q_A <- as.vector(predict(Q)$pred) # obtain predictions for everyone using the treatment they actually received
W_A1 <- W_A %>% mutate(A = 1)  # data set where everyone received treatment
Q_1 <- as.vector(predict(Q, newdata = W_A1)$pred) # predict on that everyone-exposed data set
W_A0 <- W_A %>% mutate(A = 0) # data set where no one received treatment, everyone is control
Q_0 <- as.vector(predict(Q, newdata = W_A0)$pred)
```
MAKE A 3 COLUMN DATASET WITH THE PREDICTIONS
```{r}
dat_tmle <- tibble(Y = dat_obs$Y, A = dat_obs$A, Q_A, Q_0, Q_1)
dat_tmle %>%
  head() %>%
  gt() %>%
  tab_header("TMLE Algorithm after Step 1")
```

## Straight prediction
```{r}
ate_gcomp <- mean(dat_tmle$Q_1 - dat_tmle$Q_0)
ate_gcomp
```

##   Estimate the Probability of Treatment
```{r}
A <- dat_obs$A
W <- dat_obs %>% select(-Y, -A) # matrix of predictors that only contains the confounders W1, W2, W3, and W4
g <- SuperLearner(Y = A, # outcome is the A (treatment) vector
        X = W, # W is a matrix of predictors
        family=binomial(), # treatment is a binomial outcome
        SL.library=sl_libs) # using same candidate learners 
## could use different learners
```

## The inverse probability of receiving treatment.
```{r}
g_w <- as.vector(predict(g)$pred) # Pr(A=1|W)
H_1 <- 1/g_w

## The negative inverse probability of not receiving treatment.

H_0 <- -1/(1-g_w) # Pr(A=0|W) is 1-Pr(A=1|W)

```
If the observation was treated, the inverse probability of receiving treatment, and if they were not treated, the negative inverse probability of not receiving treatment. 

```{r}
dat_tmle <- # add clever covariate data to dat_tmle
  dat_tmle %>%
  bind_cols(
         H_1 = H_1,
         H_0 = H_0) %>%
  mutate(H_A=case_when(A==1 ~ H_1,# if A is 1 (treated), assign H_1
         A == 0 ~ H_0))  # if A is 0 (not treated), assign H_0

dat_tmle %>% head() %>% gt() %>% tab_header("TMLE Algorithm after Step 2") %>% fmt_number(everything(), decimals = 2)
```
```{r}
glm_fit <- glm(Y ~ -1 + offset(qlogis(Q_A)) + H_A, data=dat_tmle, family=binomial)
eps <- coef(glm_fit)

```

```{r}

H_A <- dat_tmle$H_A # for cleaner code in Q_A_update
Q_A_update <- plogis(qlogis(Q_A) + eps*H_A)
Q_1_update <- plogis(qlogis(Q_1) + eps*H_1)
Q_0_update <- plogis(qlogis(Q_0) + eps*H_0)
```

```{r}
tmle_ate <- mean(Q_1_update - Q_0_update)
tmle_ate

```


```{r}
infl_fn <- (Y - Q_A_update) * H_A + Q_1_update - Q_0_update - tmle_ate

tmle_se <- sqrt(var(infl_fn)/nrow(dat_obs))
```


```{r}
conf_low <- tmle_ate - 1.96*tmle_se
conf_high <- tmle_ate + 1.96*tmle_se
pval <- 2 * (1 - pnorm(abs(tmle_ate / tmle_se)))

tibble(tmle_ate, conf_low, conf_high) %>% head() %>% gt() %>% tab_header("TMLE Estimate of the ATE") %>% fmt_number(everything(), decimals = 3)

```
```{r}
install.packages("tmle")
tmle_fit <-
  tmle::tmle(Y = Y, # outcome vector
           A = A, # treatment vector
           W = W, # matrix of confounders W1, W2, W3, W4
           Q.SL.library = sl_libs[1:3], # superlearning libraries from earlier for outcome regression Q(A,W)
           g.SL.library = sl_libs) # superlearning libraries from earlier for treatment regression g(W)

tmle_fit
```

