---
title: "Workshop - Day 1"
author: "Rudradev Sengupta"
date: "2/16/2024"
output: 
  html_document:
    theme: cerulean
    toc: true
    toc_depth: 3
    toc_float: true
    collapsed: true
    code_folding: hide
---

# Introduction and basics

## Load packages

```{r setup, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(MASS)
library(tidyverse)
library(gridExtra)

```

## Example: Normal Distribution

### Displayed by Column

```{r normaldist}

p1 <- ggplot(data.frame(x = c(-6, 6)), aes(x = x)) +
  stat_function(fun = dnorm, args = c(mean =0, sd =1)) + theme_bw()


p2 <- ggplot(data.frame(x = c(-6, 6)), aes(x = x)) +
  stat_function(fun = dnorm, args = c(mean =1, sd =1)) + theme_bw()

gridExtra::grid.arrange(p1,p2, ncol = 2)

```


### Displayed by Row

```{r normaldist2}

p3 <- ggplot(data.frame(x = c(-6, 6)), aes(x = x)) +
  stat_function(fun = dnorm, args = c(mean =0, sd =4)) + theme_bw()

gridExtra::grid.arrange(p1,p3, ncol = 1)

```


# Dataset (cars and diamonds)


## Display using head()

```{r cars}

## display part of the data
head(mtcars)
head(diamonds)
```

## Display using tibble()

```{r}
## display as a tibble
tibble(mtcars)
tibble(diamonds)
```

## Exploratory Scatterplot

### Simple Scatterplot

```{r}
plot(mpg~wt, data=mtcars, pch=19)
```

### Scatterplot with ggplot2
```{r}
ggplot(mtcars, aes(wt,mpg)) + geom_point()
```

### Can we make it better ?

```{r}
## parameters to control the aesthetics
title_fontsize <- 15
subtitle_fontsize <- 12
axisText_fontsize <- 12

ggplot(mtcars, aes(wt,mpg)) + geom_point(col="steelblue", size=3) + 
labs(title="mtcars dataset", subtitle="mpg vs wt", y="Miles/Gallon", x="Weight", caption="Simple Scatterplot") + 
theme_bw()+
  theme(axis.title = element_text(size = title_fontsize, face = "bold"),
        axis.text.y = element_text(size = axisText_fontsize, face = "bold"),
        axis.text.x = element_text(size = axisText_fontsize, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = title_fontsize, face = "bold"),
        plot.subtitle = element_text(size = subtitle_fontsize, face = "bold"),
        plot.caption = element_text(size = title_fontsize, face = "bold")    
)

```


## Boxplot

### Simple boxplot 
```{r}

plot(mpg~factor(vs), data=mtcars)
```

### Boxplot using ggplot
```{r}
ggplot(mtcars, aes(x=factor(vs), y=mpg)) +
  geom_boxplot() +
    labs(title="mtcars dataset", subtitle="mpg vs factor(vs)", y="Miles/Gallon", x="Not V-shaped", caption="Simple Boxplot") + 
  theme_bw()+
  theme(axis.title = element_text(size = title_fontsize, face = "bold"),
        axis.text.y = element_text(size = axisText_fontsize, face = "bold"),
        axis.text.x = element_text(size = axisText_fontsize, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = title_fontsize, face = "bold"),
        plot.subtitle = element_text(size = subtitle_fontsize, face = "bold"),
        plot.caption = element_text(size = title_fontsize, face = "bold"))   

```


# Linear Regression

## Model using lm() and visualization with ggplot
```{r}

lm.mod=lm(mpg~wt, data=mtcars)

## display results
summary(lm.mod)
```

```{r message=FALSE}
## visualize results
ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(col="steelblue", size=3) +
  geom_smooth(method="lm", col = "black") + 
  labs(title="mtcars dataset", subtitle="mpg vs wt", y="Miles/Gallon", x="Weight", caption="Simple Scatterplot") + 
  theme_bw()+
  theme(axis.title = element_text(size = title_fontsize, face = "bold"),
        axis.text.y = element_text(size = axisText_fontsize, face = "bold"),
        axis.text.x = element_text(size = axisText_fontsize, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = title_fontsize, face = "bold"),
        plot.subtitle = element_text(size = subtitle_fontsize, face = "bold"),
        plot.caption = element_text(size = title_fontsize, face = "bold"))    

```

## Implementation with glm()

```{r message=FALSE}
lm.fit=lm(mpg~wt, mtcars)
glm.fit=glm(mpg~wt, mtcars, family="gaussian") ## add the family parameter


ggplot(mtcars, aes(x=wt, y=mpg)) +
  geom_point(col="steelblue", size=3) +
  geom_smooth(method="glm", col = "black") + 
  labs(title="mtcars dataset", subtitle="mpg vs wt", y="Miles/Gallon", x="Weight", caption="Simple Scatterplot") + 
  theme_bw()+
  theme(axis.title = element_text(size = title_fontsize, face = "bold"),
        axis.text.y = element_text(size = axisText_fontsize, face = "bold"),
        axis.text.x = element_text(size = axisText_fontsize, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = title_fontsize, face = "bold"),
        plot.subtitle = element_text(size = subtitle_fontsize, face = "bold"),
        plot.caption = element_text(size = title_fontsize, face = "bold"))   
```


# Running ANOVA

```{r}
aov.mod=aov(mpg~factor(vs), data=mtcars) 
summary(aov.mod)
```

The output answers a major question: there is a significant difference in fuel efficiency based on engine shape.

## ANOVA as a Linear Model, which is a GLM

ANOVAs are just a type of linear model in which the predictor variable is categorical. This means that we can actually run the ANOVA using the lm() function as well! However, there is a slight difference in how to get the results of test.

```{r}
lm.mod2=lm(mpg~factor(vs), data=mtcars) 
lm.mod2
summary(lm.mod2)
```

Notice that the test statistic is a *t-value* instead of *F-value* that we got when we ran the aov() function. The results are identical when we run the following.

```{r}
anova(lm.mod2)
```

````{r}
glm.mod2=glm(mpg~factor(vs), data=mtcars, family="gaussian") 
anova(glm.mod2, test="F")
```

## Multiple predictor variables in a Linear Model

We will plot mpg against weight, and then use engine shape as the point color.

```{r}
legendTitle_fontsize = axisText_fontsize
legendText_fontsize = legendTitle_fontsize - 2

ggplot(mtcars, aes(x=wt, y=mpg, color=factor(vs))) +
  geom_point(size = 3) +
  scale_color_discrete(name="Engine Shape",labels=c("v-shape", "straight")) +
    labs(title="mtcars dataset", subtitle="mpg vs wt / factor(vs)", y="Miles/Gallon", x="Weight", caption="Simple Scatterplot by Group") + 
  theme_bw()+
  theme(axis.title = element_text(size = title_fontsize, face = "bold"),
        axis.text.y = element_text(size = axisText_fontsize, face = "bold"),
        axis.text.x = element_text(size = axisText_fontsize, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = title_fontsize, face = "bold"),
        plot.subtitle = element_text(size = subtitle_fontsize, face = "bold"),
        plot.caption = element_text(size = title_fontsize, face = "bold"),
        legend.title = element_text(face = "bold", size=legendTitle_fontsize),
        legend.text = element_text(face = "bold", size=legendText_fontsize))   
```


We can now actually add the fit lines using *geom_smooth()* on top of this, and you will get two lines: one for each engine shape.


```{r}
ggplot(mtcars, aes(x=wt, y=mpg, color=factor(vs))) +
  geom_point(size = 3) +
  scale_color_discrete(name="Engine Shape",labels=c("v-shape", "straight")) +
    labs(title="mtcars dataset", subtitle="mpg vs wt / factor(vs)", y="Miles/Gallon", x="Weight", caption="Simple Linear Regression by Group") + 
  geom_smooth(method="lm")+
  theme_bw()+
  theme(axis.title = element_text(size = title_fontsize, face = "bold"),
        axis.text.y = element_text(size = axisText_fontsize, face = "bold"),
        axis.text.x = element_text(size = axisText_fontsize, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = title_fontsize, face = "bold"),
        plot.subtitle = element_text(size = subtitle_fontsize, face = "bold"),
        plot.caption = element_text(size = title_fontsize, face = "bold"),
        legend.title = element_text(face = "bold", size=legendTitle_fontsize),
        legend.text = element_text(face = "bold", size=legendText_fontsize))   

```

+ Is the difference in mpg between the engine shapes actually driven by vehicle weight?

+ Is there still an effect after accounting for engine weight?

+ Does the relationship between weight and mpg contingent on engine shape?

For the first two questions, we could run a model where we just add the two predictor variables in the formula:
```{r}
lm.mod3=lm(mpg~ wt + factor(vs), data=mtcars)
anova(lm.mod3)
```

This model output suggests that most of the variation in mpg is explained by vehicle weight, but that there is also additional effect of engine shape. This is because the F value for weight is larger (and P-value is lower) than for engine shape.

To answer the last question, we add the “interaction term” for weight and engine shape. 

```{r}

lm.mod4=lm(mpg~ wt * factor(vs), data=mtcars)
anova(lm.mod4)

```

Output suggests that there is a significant interaction between vehicle weight and engine shape on their effect on fuel efficiency. This approach is called ANCOVA.

# GLMs with binary, count and proportion data as response variable



## Binary data

### Basic visualization of the diamonds dataset

```{r}
ggplot(diamonds, aes(x = cut , fill = cut)) +
  theme_bw() +
  geom_bar()+
  labs(x = "Quality of Diamonds",
       y = "Diamonds Count",
       title = "Quality of the Diamonds")


ggplot(diamonds, aes(x = carat , y = price)) + geom_point(size = 3) +
  theme_bw() +
  labs(x = "Weight of Diamonds",
       y = "Price of Diamonds",
       title = "diamonds dataset")

ggplot(diamonds, aes(x = carat , y = price, color = cut)) + geom_point(size = 3) +
  theme_bw() +
  labs(x = "Weight of Diamonds",
       y = "Price of Diamonds",
       title = "diamonds dataset")
```

Let’s say we want to predict when a diamond will cost above a certain threshold price (let’s say \$10,000). To do this, let’s make a new variable called “expensive”, which is TRUE if the diamond is priced at or above \$10,000, and FALSE if not.

```{r}

diamonds$expensive= ifelse(diamonds$price >= 10000,1,0)

```

Now, we can test if the probability that a diamond is worth > \$10,000 is dependent on its “carat”.

```{r}

glm.fit=glm(expensive~carat, data=diamonds)
summary(glm.fit)

```

If we want an F statistic:
```{r message=FALSE}
anova(glm.fit, test="F")
```


Visualize the model: 

```{r message=FALSE}

ggplot(diamonds, aes(x=carat, y=expensive))+
  geom_point() +
  geom_smooth(method="glm", method.args = list(family = "binomial")) + 
      labs(title="diamonds dataset", subtitle="carat vs expensive", 
           y="Expensive", x="Carat", caption="Logistic Regression") + 
  theme_bw()+
  theme(axis.title = element_text(size = title_fontsize, face = "bold"),
        axis.text.y = element_text(size = axisText_fontsize, face = "bold"),
        axis.text.x = element_text(size = axisText_fontsize, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = title_fontsize, face = "bold"),
        plot.subtitle = element_text(size = subtitle_fontsize, face = "bold"),
        plot.caption = element_text(size = title_fontsize, face = "bold")) 

```


## Count and proportional data

### Count data

We fit a Poisson regression model using the count of deaths as the response variable from the *snails* dataset. The analysis is restricted to snails held for 3 or 4 weeks because we expect the effect to be less apparent if the snails are kept in the experiment for a short amount of time.

```{r}
count.fit=glm(Deaths~Temp, data=snails, subset=which(Exposure>=3), family="poisson")
summary(count.fit)

```

Visualize the results:

```{r message=FALSE}

ggplot(snails[which(snails$Exposure>=3),], aes(x=Temp, Deaths)) +
  geom_count() +
  geom_smooth(method="glm", method.args=list(family="poisson")) + 
      labs(title="snails dataset", subtitle="Temp vs Deaths", y="Deaths", x="Temperature (Celsius)", caption="Poisson Regression by Group") + 
  theme_bw()+
  theme(axis.title = element_text(size = title_fontsize, face = "bold"),
        axis.text.y = element_text(size = axisText_fontsize, face = "bold"),
        axis.text.x = element_text(size = axisText_fontsize, face = "bold"),
        panel.grid = element_blank(),
        plot.title = element_text(size = title_fontsize, face = "bold"),
        plot.subtitle = element_text(size = subtitle_fontsize, face = "bold"),
        plot.caption = element_text(size = title_fontsize, face = "bold"),
        legend.title = element_text(face = "bold", size=legendTitle_fontsize),
        legend.text = element_text(face = "bold", size=legendText_fontsize)) 

```


### Proportion data

```{r}
prop.fit=glm(cbind(Deaths, N)~Temp, data=snails, subset=which(Exposure>=3), family=binomial(link="logit"))
summary(prop.fit)

```

