---
title:  "Introduction to R: Simple linear regression"
author:
- name: Ziv Shkedy
  affiliation: Hasselt University, Belgium
subtitle:  First steps in statistical modeling in R

#institute: UHasselt
date: May, 2020
output:
  rmdformats::readthedown:
    highlight: kate
    use_bookdown: TRUE
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
categories: ["R"]
bibliography: bibliography.bib
---


<script type="text/javascript"
  src="http://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML">
</script>


```{r style, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
suppressPackageStartupMessages({
  library(BiocStyle)
  library(knitr)
  library(rmarkdown)
  library(bookdown)
})
options(width = 100)
opts_chunk$set(fig.width = 5, fig.height = 5)

```

```{r,echo=FALSE}
library(e1071)
library(lattice)
library(ggplot2)
library(mvtnorm)
```

# Introduction

## Slides, code and tutorials

This part of the interactive book contains all R code that was used to produce the results and output presented in chapter 3 (modeling: simple linear regression) in the course's slides. We include YouTube tutorials as a part of the book and links to the relevant tutorials are provided in different chapters. Note that these tutorials were not developed especially for this book, they cover the same topics using different examples. 

## R ?
No previous knowledge about R is required. We use the R function <tt> lm() </tt> to a fit simple linear regression model R and the cars dataset <tt> cars </tt> is used for illustraion. The model can be fitted using the function <tt> glm() </tt> as well.

##  Slides
Slide for this part of the course are avilable online in the >eR-BioStat website. See  [RcourseModeling](https://github.com/eR-Biostat/Courses/blob/master/Introductory%20Courses/Introduction%20to%20R/Slides/eR-Biostat_An%20Introduction%20to%20R_2017_Modeling1.pdf).

# The cars data

The cars data gives the speed of cars (the response) and the distances taken to stop (the predictor). Note that the data were recorded in the 1920s.

```{r}
x<-cars$speed
y<-cars$dist
head(cars)
```

# Fitting Simple Linear regression Model in R


## YouTube tutorials: Simple linear regression in R

### R - Simple Linear Regression (part 1)

For a YouTube tutorial about simple linear regression in R by Jalayer Academy see [YTREgression1a](https://www.youtube.com/watch?v=wnIlld_8lSg).

### R - Simple Linear Regression (part 2)

For a YouTube tutorial about simple linear regression by Jalayer Academy in R see [YTREgression1b](https://www.youtube.com/watch?v=m_8XYVzT1IU).

### Simple linear regression in R | R Tutorial 5.1

For a YouTube tutorial about simple linear regression in R by MarinStatsLectures see [YTREgression2](https://www.youtube.com/watch?v=66z_MRwtFJM&t=14s).


##  The <tt> lm() </tt> Function

The R function which we use to fit a linear regression model is <tt> lm() </tt>. A General call of the function has the form of <tt> lm(dependent variable~predictor) </tt>.

## Scatterplot

Figure \@ref(fig:fig1) shows the scatterplot of Y versus X.

```{r fig1,fig.cap="Speed versus stopping ditance."}
plot(x, y)
title("Y vs. X")
```


## Fitting the regression model in R

In order to fit a simplie linear regression model, 
\[
Y_{i}=\beta_{0}+\beta_{1}X_{i}+\varepsilon_{i},\;\;i=1,\dots,n,
\]

we use the <tt> lm() </tt> function in the following way:

```{r}
fit.LM <- lm(y ~ x)
```

The object <tt> fit.LM </tt>  contains the results of the estimated model.

```{r}
fit.LM
```

## Parameter estimates: inference

Parameter estimates, standard errors, t-tests (and p-values) are obtained using the function <tt>  summary() </tt>

```{r}
summary(fit.LM)
```

We use the function <tt> anova() </tt> in order to obtain the ANOVA Table and F-test.

```{r}
anova(fit.LM)
```




# Residuals and model diagnostic

## Data and fitted model

The object <tt> fit.LM\$fit </tt> contains the fitted values.

```{r}
fit.LM$fit
```


Figure \@ref(fig:fig2)shows the scaterplot of Y versus X and the fitted values.

```{r fig2,fig.cap="Data and fitted model."}
plot(x, y)
lines(x, fit.LM$fit)
title("data and fitted model")
```

Note that the straight (the fitted model) line is

\[
\hat{Y}_{i}=\hat{\beta}_{0}+\hat{\beta}_{1}X_{i}.
\]

## Distribution of the residuals

The object <tt> fit.LM\$resid </tt> contains the residuals: $e_{i}=Y_{i}-\hat{Y}_{i}$.

```{r}
fit.LM$resid
```

Histogram and normal probability plot for the residuals are shown in Figure \@ref(fig:fig3).

```{r  fig3,fig.cap="Distribution of the residuals."}
par(mfrow = c(1, 2))
hist(fit.LM$resid)
qqnorm(fit.LM$resid)
```

Note that the set of diagnostic plots can be produced using the function <tt>plot() </tt> with the object <tt> fit.LM </tt>. 

```{r fig4,fig.cap="Diagnostic plots."}
plot(fit.LM)
```



