---
title:  "Introduction to R: One-Way ANOVA "
author:
- name: Ziv Shkedy
  affiliation: Hasselt University, Belgium
subtitle:  First steps in statistical modeling in R (July 2020)

#institute: UHasselt
date: July, 2020
output:
  rmdformats::readthedown:
    highlight: kate
    use_bookdown: TRUE
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
categories: ["R"]
bibliography: bibliography.bib
---


<script type="text/javascript"
  src="http://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML">
</script>


```{r style, echo=FALSE, message=FALSE, warning=FALSE, results="asis"}
suppressPackageStartupMessages({
  library(BiocStyle)
  library(knitr)
  library(rmarkdown)
  library(bookdown)
})
options(width = 100)
opts_chunk$set(fig.width = 5, fig.height = 5)

```

```{r,echo=FALSE}
library(e1071)
library(lattice)
library(ggplot2)
library(mvtnorm)
```

# Introduction

## Slides, code and tutorials

This chapter of the interactive book contains all R code that was used to produce the results and output presented in chapter 3 (modeling :one way ANOVA) in the course's slides. We include YouTube tutorials as a part of the book chapter and links to the relevant tutorials are provided. Note that these tutorials were not developed especially for this book, they cover the same topics using different examples. 

## R ?
No previous knowledge about R is required. We use the R function <tt> aov() </tt> to fit a One-Way ANOVA in R and the chicken Weights dataset is used for illustraion. The same model can be fitted using the functions <tt> lm() </tt> and <tt> glm() </tt> as well.

##  Slides
Slide for this part of the course are avilable online in the >eR-BioStat website. See  [Rcoursemodeling](https://github.com/eR-Biostat/Courses/blob/master/Introductory%20Courses/Introduction%20to%20R/Slides/eR-Biostat_An%20Introduction%20to%20R_2017_Modeling1.pdf).

# The Chicken Weights data
For illustrations, we use the Chicken Weights data. Newly hatched chicks were randomly allocated into six groups, and each group was given a different feed supplement. Their weights (the response variable) in grams after six weeks are given along with feed types (the factor). The Chicken Weights data is a data frame in R called <tt> chickwts </tt> is shown below. 

```{r}
head(chickwts)
```

# One-Way ANOVA model


## YouTube tutorials: One-Way ANOVA in R

### R - One-way ANOVA

For a YouTube tutorial about One-Way ANOVA in R R Statistics and Research see [YTOneWayANOVA1](https://www.youtube.com/watch?v=4DeCaCaC2JQ).

### One way ANOVA in RStudio

For a YouTube tutorial about One-Way ANOVA by 
Tom Sherrattin R see [YTOneWayANOVA2](https://www.youtube.com/watch?v=zDGQxC0bWn4).

## Model formulation

We consider a one-way ANOVA model for that data

\[
Y_{ij}=\mu_{i}+\varepsilon_{ij}\;\;\;\;i=1,\dots,6 ,\;j=1,\dots,n_{i}. 
\]
Here,

 * $Y_{ij}$ is the weight of the $j$'th subject in diet group $i$.
 * The parameters $\mu_{i}$ represent the mean of the distribution of weight  at each age group.
 * \item $\varepsilon_{ij}$ is a random error which assumed to be normally distributed,      
   $$\varepsilon_{ij}\sim\mbox{N}(0,\sigma^{2})$$.
   

Our primary of interest is to estimate the group means and then to test the hypotheses
\[
\begin{array}{l}
\mbox{H}_{0}:\mu_{1}=\mu_{2}=\dots=\mu_{6},\\
\mbox{H}_{1}: \mu_{l} \ne \mu_{k}\;\;\mbox{for at least one pair}.
\end{array}
\]

## Visualizing the Data 

The striptplot in Figure~\@ref(fig:fig1) shows the chickens' weight by diet group. 

```{r fig1, fig.cap="Chicks weight by diet group."}
stripplot(feed ~ jitter(weight), 
            data = chickwts, 
            aspect = 1, jitter = T, 
            xlab="Weight", col = 1)
```

The boxplot in Figure~\@ref(fig:fig2) can be used to visualize the patterns in the data. Note how the box of the horsebean diet group located lower than the boxes of the other diet groups.

```{r fig2, fig.cap="Chicks weight by diet group."}
boxplot(split(chickwts$weight,chickwts$feed))
```

The sample means for the 6 diet groups are equal to

```{r}
 tapply(chickwts$weight,chickwts$feed,mean)
```

# Fitting a One-way ANOVA Model in R

## The <tt> aov() </tt> Function

The R  function which we use to fit a One-way ANOVA  model in R is <tt>  aov() </tt>. A General call of the function has the form of <tt> aov(dependent variable~factor) </tt>.  For example, a one-way ANOVA model for the response <tt> y </tt> and the factor <tt> x </tt> can be fitted using <tt> aov(y~x) </tt>. If <tt> x </tt> is a numerical vector, we can use <tt> aov(y~as.factor(x)) </tt>.

##  Fitting an ANOVA Modelfor the Chicken Weights data

 In order to fit the model 
\[
Y_{ij}=\mu_{i}+\varepsilon_{ij},
\]

we use

```{r}
Fit.aov<-aov(chickwts$weight~chickwts$feed)
```
the object <tt> Fit.aov </tt> contains the results. The ANOVA table can be produced using the function <tt> summary() </tt> and it is given by:

```{r}
 summary(Fit.aov)
```


# Diagnostic Plots

## Residuals

The object <tt> Fit.aov$resid </tt> contains the residuals values.

```{r}
Fit.aov$resid
```

A Stem-and-leaf diagram reveals a symatical distribution of the residuals.

```{r}
stem(Fit.aov$resid)
```

Stripplot for the residuals by diet group is shown in Figure~\@ref(fig:fig3)

```{r fi3, fig.cap="Residuals by diet group."}
stripplot(chickwts$feed ~ jitter(Fit.aov$resid), aspect = 1, jitter = T, xlab = 
	"Residuals", col = 1)
```

Boxplot, histogram and normal probability plot for the residuals are shown in Figure ~\@ref(fig:fig4).

```{r fig4,fig.cap="Distribution for the residuals."}
par(mfrow = c(2, 2))
hist(Fit.aov$resid)
boxplot(split(Fit.aov$resid, chickwts$feed))
qqnorm(Fit.aov$resid)
```

Alternativly, these figures can be produce using the function <tt> plot() </tt> in the following way

```{r fig5, fig.cap="Diagnostic plots."}
plot(Fit.aov)
```





