---
title: "The tidyverse"
subtitle: " "
author: "Ziv SHkedy et al. (2020)"
institute: "Hasselt University"
date: "16/09/2020"
output:
  xaringan::moon_reader:
    css: ["default", "chocolate", "rladies-fonts"]
    lib_dir: libs
    seal: false
    nature:
      beforeInit: "cols_macro.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,	
                      warning = FALSE,
                      eval = TRUE,
                      tidy = FALSE)
library(knitr)
library(tidyverse)
library(deSolve)
library(minpack.lm)
library(ggpubr)
library(readxl)
library(gamlss)
library(data.table)
library(grid)
library(png)
library(nlme)
library(gridExtra)
library(mvtnorm)
library(e1071)
library(lattice)
library(ggplot2)
library(dslabs)
library(NHANES)
library(dplyr)

```



class: center, middle


# An introduction to the <tt>tidyverse</tt>

#### Ziv Shkedy et al. (2020)


---


class: center, middle

# Part 1: Introduction 

`Focus: a quick introduction to tidyverse` <br>

---


# The <tt> tidyverse </tt>

 * The <tt> tidyverse </tt> is a collection of R packages that work in data frames in a tidy format.
 * All packages are uploaded in CRAN, and can be installed  using <tt>  install.packages </tt> . 
 * In this chapter of the interactive book, we cover materials at an introduction level and follow closely the topics    
   presented in Chapter 4 in the book  Data Analysis and Prediction Algorithms with R by 
   Rafael A. Irizarry.

---

# What do we cover in this chapter ?

* The chapter is developed at a beginner level.
* We cover few functions from the <tt> tidyverse </tt> packages and illustrate the basic concepts using different examples   of the following functions:

 - <tt> mutate() </tt>
 - <tt> filter() </tt>
 - <tt> select() </tt> (separate set of slides)
 - The pipe: $\% > \%$ (separate set of slides)
 - <tt> summarize() </tt>
 - <tt> group_by() </tt>
 - <tt> arrange() </tt>
 - <tt> top_n() </tt>
 
---


# <tt>ggplot2</tt> ?

* Our aim in this tutorial is not to teach <tt> ggplot2 </tt> (this will be done in a different chapter). 
* Some functions of the package are used to visualized the main patterns in the datasets we used to illustrate the examples presented in this chapter. 
* The following graphical functions are used for visualization:

  - <tt> qplot() </tt>
  - <tt> ggplot() + geom_jitter()</tt>
  - <tt> ggplot() + geom_point() </tt>
  - <tt> gplot() +    geom_density() </tt>
  - <tt>stripplot()</tt>


---

# Online references (I)

* Materials about <tt> tidyverse </tt> are widely available online. 

 * For a YouTube tutorial about tidyverse in R by 
   Mark Gingrass see [YTtidyverse1](https://www.youtube.com/watch?v=JtQfXY0lIzc).
 * For a YouTube tutorial about tidyverse in R by 
   Garreet Grolemund see [YTtidyverse2a](https://www.youtube.com/watch?v=jOd65mR1zfw).
 * For a YouTube tutorial about tidyverse in R by 
   Garreet Grolemund see [YTtidyverse2b](https://www.youtube.com/watch?v=1ELALQlO-yM).
 * For a YouTube tutorial about tidyverse in R by 
   Ben Stenhaug see [YTtidyverse3](https://www.youtube.com/watch?v=lTTJPRwnONE&list=PLLxj8fULvXwGOf8uHlL4Tr62oXSB5k_in).

---

# Online references (II)

* Interactive book for the course: [Booktidyverse1](http://htmlpreview.github.io/?https://github.com/eR-Biostat/Courses/blob/master/Introductory%20Courses/Introduction%20to%20R/Onlinebook/Rintro_Tidy-html-_V1.html)
* Chapter 4 in the book: Data Analysis and Prediction Algorithms with R by Rafael A. Irizarry see [Chapter4](https://rafalab.github.io/dsbook/tidyverse.html).

---

class: center, middle

# Part 2: the tidyverse package and tidy data

`Focus: modern data frames` <br>


---

# Tidy data

* Tidy data is a data format in which each row represents one measurement for one observation and columns are, as usual, the variables in the data.

---


# The murders data

.pull-left[
* The <tt>murders</tt> is an example of a tidy data. 
* The information for each state is given in one line. 
* Note that in this case, each observation (=state) has information in one data line.
]

.pull-right[
```{r}
data("murders")
head(murders) 
```
]

---

# The murders data
.pull-left[
* The murder rate by region  is shown in the stripplot.
* It shows clearly that in the west, the total murders is the lowest.
]

.pull-right[
```{r fig1b,fig.cap="Dotplot using the gg2plot package"}
ggplot(murders, aes(region,total)) + geom_jitter(position = position_jitter(width = .05))
```
]

---

# The ChickWeight data

* The Chicken Weight data gives information about weights of chicken in differnt diet groups.
* Each observation is a chick and it was measured in 12 times points. 


```{r}
names(ChickWeight)
unique(ChickWeight$Time)
```

* In the data frame, each measurement is presented in one data line. 
* Hence, the <tt> ChickWeight </tt> is a tidy data. 

---

# The ChickWeight data
.pull-left[
* This implies that that the data for each observation is presented in 12 lines.
* note that NOT ALL subjects were measured in 12 times. 
* Data for the first 6 time points of the first chick are listed below.
]

.pull-right[
```{r}
head(ChickWeight)
```
]

---

# The ChickWeight data

.pull-left[
* The boxplot of the chicken weights by time point.
* Itshows the increasing trend of the weight over time.
]

.pull-right[
```{r fig1c,fig.cap="Boxplot for the singers data (I.1)."}
ggplot(ChickWeight, aes(as.factor(Time),weight)) + geom_boxplot()
```
]

---

class: center, middle

# Part 3: the <tt>mutate()</tt> 

`Focus: adding a column(s)` <br>

---

#  Adding a variable (column)

* Suppose that we would like to calculate the murder rate per 100000 people  that is

$$rate=\frac{total}{population} \times 100000.$$
* This can be done using the <tt> mutate() </tt> funtion.
* The function has the grneral call of:  <tt> mutate(data frame , new variable) </tt>.

---

# The murders data

* For the <tt>murders</tt> data we have

```{r}
data("murders")
murders <- mutate(murders, rate = total / population * 100000)
```

* Note that after calculating the murder rate, the <tt> murders </tt> has an extra column (=variable) for the <tt>rate</tt>.

```{r}
head(murders) 
```

---

# The NHANES data

* The BMI of a person is given by $$BMI=\frac{weight}{height^{2}}$$.


* To calculate the BMI in the <tt> NHANES </tt> we use

```{r}
data("NHANES")
Data_new <- mutate(NHANES, BMI_new = Weight / (Height*Height))
```

---

# The NHANES data

.pull-left[
* The histogram of the BMI.
]

.pull-right[
```{r fig1d,fig.cap="Histogram for the BMI."}
qplot(BMI_new , data=Data_new, geom="histogram")
```
]
---

# Practical session

* For the cars data (<tt> mtcars</tt>):
  
  - Define a new variable: mile per gallon / weight.
  - Produce a boxplot for the variable mile per gallon / weight by number of number of cylinders.

* For the murders data :
  
  - Produce a boxplot for the murder rate by region.

---

class: center, middle

# Part 4: the <tt> filter() function </tt> 

`Focus: selection of cases` <br>

---

# Filtering

* Selection of observations (=cases) from the data.
* A subset of the data.
* can be done using the function <tt> filter() </tt>.
* The new data: has the same variables.

---

# The murders data

* For murder dataset, we can select all the states with murder rate $\le 0.71$ by

```{r}
filter(murders, rate <= 0.71)
```

---

# The cars data

* For cars data, suppose that we would like to plot the cars' weight versus the cars mpg for cars with weight $\le 3$. 
* We first select all cars with weight smaller or equal to 3:  

```{r}
dim(mtcars)
mtcars1<-filter(mtcars, wt <= 3)
dim(mtcars1)
```

---

# The cars data

* The new data frame <tt> mtcars1 </tt> contains the information for all cars with weight lower than 3,

```{r}
mtcars1
```
---

# The cars data

.pull-left[
* Car's weight <= 3.
* The scaterplot below of the weight versus the mpg can be produce using the following code.
]

.pull-right[
```{r fig1e,fig.cap="mile per galllon versus weight."}
ggplot(mtcars1, aes(x=wt, y=mpg)) + 
  geom_point()
```
]

---

# Practical session

* For the NHANES data:

 - Select all the observations for which the BMI is greater or equal to 30.
 - Produce the histogram for the variable BMI for all male. 

* For the chick weight data (<tt>chickwts</tt>):

 - Select all chicks with weight smaller than 158.
 - Select all chicks from the <tt>horsebean</tt> feed group.

---


class: center, middle

# Part 5: the <tt>select()</tt> function 

`Focus: selection of variables` <br>

---


class: center, middle

# Part 6: the pipe $\% > \%$ 

`Focus: multiple startments in one run` <br>

---


class: center, middle

# Part 7: the <tt> summarize()</tt> function 

`Focus: summary statistics` <br>

---

# The summarize function 

* The function <tt> summarize() </tt> allows us to produce summery statistics for variables in the data frame.

---

# The heights data

* The height data frame gives the height by gender for 1050 individuals

```{r}
library(dplyr)
library(dslabs)
data(heights)
dim(heights)
```

* The first 6 observations in the data are listed below.

```{r}
head(heights)
```

---

## Mean and standard deviation

* We can calculate the mean and standard deviation for female using the function <tt> summarize () </tt>.
* Note that we first filter the data and define and subgroup contains the data for female

```{r}
s <- heights %>%  filter(sex == "Female") %>%
  summarize(average = mean(height), standard_deviation = sd(height))
```

* The object <tt> s </tt> stores the results

```{r}
s
```

---

## Mean and standard deviation

* We can define a vector that contains the results


```{r}
c(s$average,s$standard_deviation)
```

---

# Summary statistics by group

Alternatively, we can define a vector with the female heights (<tt> height.female </tt> ) and calculate the mean and standard deviation for this vector.

```{r}
height.female<-heights$height[heights$sex == "Female"]
mean(height.female)
sd(height.female)
```

---

## Quantiles

* The median, minimum and maximum height for female

```{r}
heights %>% 
  filter(sex == "Female") %>% summarize(median = median(height), minimum = min(height), 
                                         maximum = max(height))
```

---

## Quantiles

* These summary statistics can be also calculate can the function <tt> quantile </tt>.

```{r}
heights %>%  filter(sex == "Female") %>%
  summarize(range = quantile(height, c(0, 0.5, 1)))
```


---

# The chicks data

* To calculate the mean and standard deviation for the chick weights we use

```{r}
s <- chickwts %>%  summarize(average = mean(weight), standard_deviation = sd(weight))
s
```

* Note that for this example we ignore the diet group. 

---

# Practical session

* For the NHANES data:
  
  * Canculate the mean, meadian and standard deviation of the variable BMI.
  * Canculate the mean, meadian and standard deviation of the variable BMI only for male.

* For the cars data:
  
  * Canculate the mean, meadian and standard deviation of the variable weight.
  * Canculate the mean, meadian and standard deviation of the variable weight only for cars with       
    automatic transmission.
  
---


class: center, middle

# Part 8: the <tt> group_by()</tt> function 

`Focus: summary statistics by group` <br>

---

# Analysis by group

* In this section we focus on an analysis in which the analysis is done across a level of a factor in the data frame. 
* For example, the diet group in the chick data frame etc.

---


# The heights data

* The mean and standard deviation for the height by gender can be calculate by adding the function <tt> group_by(sex) </tt>

```{r}
heights %>% 
  group_by(sex) %>%
  summarize(average = mean(height), standard_deviation = sd(height))
```

---

# The heights data

* The same results can be obtained using the function <tt> tapply </tt>.


```{r}
tapply(heights$height,heights$sex,mean)
tapply(heights$height,heights$sex,sd)
```

---

# The murders data

* The median murder rate by region using the <tt> group_by(region) </tt> and the <tt> summarize () </tt> functions

```{r}
murders %>% group_by(region) %>%
  summarize(median_rate = median(rate))
```

---

# The murders data

The median murder rate by region using the <tt> tapply() </tt> function.

```{r}
tapply(murders$rate,murders$region,median)
```

---

# The chicks data

* Summary statistics by diet group

```{r}
chickwts %>% 
  group_by(feed) %>%
  summarize(average = mean(weight), standard_deviation = sd(weight))
```
---

# The chicks data

.pull-left[
* The striptplot reveals that the weights in the horsebean group are, in general, the smallest in the sample. 
]

.pull-right[

```{r figh, fig.cap=" by diet group."}
stripplot(feed ~ jitter(weight), 
            data = chickwts, 
            aspect = 1, jitter = T, 
            xlab="Chicks weight", col = 1)
```
]

---


# Practical session

* For the NHANES data:

  * Calculate the mean BMI by gender.
  * Calculate the mean age by diabetes status (the variable <tt>Diabetes</tt>).
  
* For the cars data:
  
  * Canculate the mean, meadian and standard deviation of the variable weight by transmission type (automatic vs. manual).
  * Canculate the mean, meadian and standard deviation of the variable miles per gallon by number ofcylinders.
  
---



class: center, middle

# Part 9: the <tt> arrange()</tt> function 

`Focus: sorting the data` <br>

---


# The murders data

* We can sort a data frame by a variable <tt> x </tt> using the function <tt> arrange(x) </tt>. 
* For the murder data frame, we sort the data by the population size

```{r}
murders %>%
  arrange(population) %>% head()
```

---

# The murders data

* The same sorting can be implemented using the function <tt> order () </tt>. 
* In this case the rows will be presented in the order of the population.

```{r}
data1<-murders[order(murders$population),]
head(data1)
```

---

# Sorting and changing the order

* We sort the data frame by rate from the lowest to the highest rate

```{r}
murders %>% 
  arrange(rate) %>%
  head()
```

---

# Sorting and changing the order


* We can change the order using the function <tt> desc() </tt> so the data are presented from the highest to the lowest rate in a decreasing order.

```{r}
murders %>% 
  arrange(desc(rate)) %>%
  head
```

---

# The Chicken Weight data

* The first 6 observations in the chicken weight data belongs to the first chick at time point 0 to 10.

```{r}
head(ChickWeight)
```

---

# The Chicken Weight data

* We sort the data frame according to the <tt> Time </tt> variable. After sorting, the first 6 lines in the data frame are the measurements for chock 1-6 at day 21. 

```{r}
ChickWeight %>% 
  arrange(desc(Time)) %>%
  head
```

---

# The Chicken Weight data

* We can reverse the order, in this case the first 6 lines are the measurements for check 1-6 at baseline (<tt> Time =0 </tt>).

```{r}
ChickWeight %>% 
  arrange(Time) %>%
  head
```

---

# Practical session

  * Sort the NHANES data according to the subjects' BMI.
  * For the NHANES data, select all subjects with diabetes (the variable <tt>Diabetes</tt>) and sort
    according to the subjects' age.
  * Sort the cars data according to the cars' mpg. 
  

---

class: center, middle

# Part 10: Nested sorting 

`Focus: sorting with two variables` <br>

---

# Nested sorting 

* Suppose that we want to present the data in an increasing order  of <tt> x </tt>  across a level of a factor <tt> y </tt>.  
* We can sort the data frame by a variable <tt> x </tt> within the factor levels using the function <tt> arrange(y,x) </tt>. 

---

# the murder data

* For the murder data frame, we sort the data by murder rate within the region

```{r}
murders %>% 
  arrange(region, rate) %>% 
  head()
```

---

# The cars data

.pull-left[
* The figure  shows that <tt>mpg</tt> as the number of cylinders decreases.
]

.pull-right[
```{r figi, fig.cap="MPG by number of cylinders."}
stripplot(cyl ~ jitter(mpg), 
            data = mtcars, 
            aspect = 1, jitter = T, 
            xlab="MPG", col = 1)
```
]

---

# The cars data

* We can sort the cars according to their mpg (in an increasing order) by the number of cylinders

```{r}
mtcars %>% 
  arrange(cyl, mpg) 
```

---

# Practical session

  * Sort the NHANES data according to the subjects' BMI by Gender.
  * Sort the NHANES data according to the subjects' age by smoking status (the variable    
    <tt>Smoke100</tt>).
  * Sort the Chicks Weights data (<tt>chickwts</tt>) according to the chicks' weight by diet group.
  

---


class: center, middle

# Part 11: The <tt>top_n()</tt> function 

`Focus: selection of the top K observations` <br>

---

# Top n

* To print the top $n$ observations according to the variable <tt> x </tt> we can use the function <tt> top_n(n, x) </tt>. 

---


# The murders data

* For the murders data, we print the top 5 states with the highest murder rate

```{r}
murders %>% top_n(5, rate)
```

---

# The cars data

.pull-left[
* The scaterplot of the cars' weight versus the cars' <tt>mpg</tt>.
* The top 4 cars, with the highest mpg are given below

```{r}
mtcars %>% top_n(4,mpg)
```
]

.pull-right[
```{r figj, fig.cap="MPG by number of cylinders."}
ggplot(mtcars, aes(x=wt, y=mpg)) + 
  geom_point()
```
]

---
# Practical session
 
 * Print the 10 observations in the NHANES data with the highest BMI.
 * Create a new data frame that contains the top 5 observations with the older age in in the NHANES   
   data. 


---

