---
title: "Visualizing Data and Exploratory Data analysis"
subtitle: " "
author: "Ziv SHkedy et al. (2020)"
institute: "Hasselt University"
date: "16/09/2020"
output:
  xaringan::moon_reader:
    css: ["default", "chocolate", "rladies-fonts"]
    lib_dir: libs
    seal: false
    nature:
      beforeInit: "cols_macro.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,	
                      warning = FALSE,
                      eval = TRUE,
                      tidy = FALSE)
library(knitr)
library(tidyverse)
library(deSolve)
library(minpack.lm)
library(ggpubr)
library(readxl)
library(gamlss)
library(data.table)
library(grid)
library(png)
library(nlme)
library(gridExtra)
library(mvtnorm)
library(e1071)
library(lattice)
library(ggplot2)
library(dslabs)
library(NHANES)
library(dplyr)

```



class: center, middle


# Visualizing Data and Exploratory Data analysis using R
## Part 1: Univariate datasets

#### Ziv Shkedy et al. (2020)


---


class: center, middle

# Part 1: Introduction 

---


#  Location, Spread and Shape in univariate data

* In the first part we focus on descriptive measures, numerical and graphical, to
characterize and visualize the features of a particular univariate distribution. 
* The following three main factors are usually used to specify a paticular distribution:
  * Location
  * Spread
  * Shape

* Each of these control different characteristics of a distribution.

---

# R datasets for illustraions

* In order to simplify the usage of slides, the data we used for illustrations are R datasets.
* We give a short description  of each data in the relevant slides.
* More details can be found with <tt>help(dataset)</tt> or (for datasets of the first part) in

  * The singers data: [singers](https://stat.ethz.ch/R-manual/R-devel/library/lattice/html/singer.html).
  * The airquality data: [airquality](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/airquality.html).
  * The cars data: [mtcars](https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars).
  * The Old Faithful Geyser Data: [oldfaithful](      https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/faithful.html)

---
  


# What do we cover in this part ?

* How to tell the story in a univariate data using R ?
* A very short introduction to <tt>ggplot2</tt>.
 
  * Layers and components of a graphical deaply.
  * How to develop a cool figure ?
  * Components of visualization.

* Patterns in Univariate datasets: location speard and shape.

---


# <tt>ggplot2</tt>  and <tt>lattice</tt> ?

 
* The following basic graphical functions are used for visualization:

   - <tt>plot()</tt>
   - <tt>lines()</tt>
   - <tt>hist()</tt>
   - <tt>boxplot()</tt>
   - <tt>qqnorm()</tt>
    
* The following <tt>lattice</tt> graphical functions are used for visualization:
    
   - <tt>dotplot()</tt>
   - <tt>histogram()</tt>
   - <tt>bwplot()</tt>
   - <tt>qqmath()</tt>
   
---

# <tt>ggplot2</tt>  and <tt>lattice</tt> ?
    
* The following <tt>ggplot2</tt> graphical functions are used for visualization:

   - <tt>ggplot()</tt>
   - <tt>geom_point()</tt>
   - <tt>theme_bw()</tt>
   - <tt>geom_smooth()</tt>
   - <tt>geom_histogram()</tt> 
   - <tt>geom_boxplot()</tt>
   - <tt>geom="density"</tt>
   - <tt>geom_violin()</tt>
   - <tt>facet_wrap()</tt>
   - <tt>qplot()</tt>
   - <tt>stat_summary()</tt>
   - <tt>geom_density_ridges() +  theme_ridges()</tt> 

---

# Online references

* Basic Skills in Visualising Data and Exploratory Data Analysis Using R - An interactive online book for the course: [BookVD](https://erbiostat.wixsite.com/vdeda1).
* Book: R Graphics Cookbook, 2nd edition by Winston Chang: [RGraphics](https://r-graphics.org/).
* Website: From Data to Viz: [Viz](https://www.data-to-viz.com/).

---

# The <tt>xaringan</tt> package

* Slides were produced using the <tt>xaringan</tt> package.

* Online book chapter about the <tt>xaringan</tt> package: [xaringan](https://bookdown.org/yihui/rmarkdown/xaringan.html).

---



class: center, middle

# Part 2: <tt>ggplot2</tt>: a short introduction

`Focus: Layers` <br>

---

# The layers of a <tt>ggplot2</tt> figure 

* A key idea behind <tt>ggplot2</tt> is that it allows to easily building up a complex plot layer by layer. 
* Each layer adds an extra level of information to the plot.
* In that way we can build sophisticated plots tailored to the problem at hand. 
---


# The <tt>mtcars</tt> data 

* The data gives information about fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).
```{r}
head(mtcars)
```

* For our example: <tt>mpg</tt>, <tt>hp</tt> and <tt>cyl</tt>.
---

# The <tt>mtcars</tt> data 


.pull-left[
* Bivariate data (miles/(US) per gallon, horsepower).
* Basic scatterplot in R: mgp Vs. hp.
* <tt>plot()</tt>.
* Our aim: fit a linear regression model for mpg with Horsepower as predictor.
]

.pull-right[

```{r}
plot(mtcars$hp,mtcars$mpg)
```
]

---


# Linear regression 

* A simple linear regression model in R.

```{r}
fit.lm<-lm(mtcars$mpg~mtcars$hp)
```

* Fitted model

```{r}
summary(fit.lm)
```

---


# First layer: basic plot 

.pull-left[
* Basic plot using <tt>ggplot2</tt>.
* Scaterplot and labels.
]

.pull-right[
```{r}
library(ggplot2)
gg <- ggplot(mtcars, aes(hp, mpg)) +
        geom_point() +
        labs(x = "Horsepower",y= "Miles Per Gallon")
gg
```
]

---



# What is <tt>aes()</tt> ?

* <tt>aes()</tt>: generate aesthetic mappings that describe how variables in the data are mapped to visual properties (Aesthetics) of geoms.

* <tt>aes()</tt> is a quoting function. This means that its inputs are quoted to be evaluated in the context of the data.

* <tt>aes(hp, mpg)</tt>: use the vraiables <tt>mpg</tt> and <tt>hp</tt> in the plot.

* ggplot2 Cheat Sheet: [ggplot2](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

---

# Second layer: add Cylinders information 

.pull-left[
* Add colors to the data points (by Cylinders number).
* <tt>aes(color=as.factor(cyl)), size=5</tt>.
]

.pull-right[
```{r}
library(ggplot2)
gg <- ggplot(mtcars, aes(hp, mpg)) +
        geom_point(aes(color=as.factor(cyl)), size=5) +
        labs(x = "Horsepower",y= "Miles Per Gallon",color= "# of Cylinders")
gg
```
]

---

# What is <tt>geom</tt> ?

* A ggplot2 <tt>geom</tt> tells the plot how do we want to display the data in R. 
* For example, <tt>geom_bar()</tt>  makes a bar chart.
* <tt>geom_point(aes(color=as.factor(cyl)), size=5)</tt>: produce a scaterplot with points.
* <tt>aes(color=as.factor(cyl))</tt>: use different colors acording to the <tt>cyl</tt> levels.
* ggplot2 Cheat Sheet: [ggplot2](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf).

---


# Third layer: change the background 


.pull-left[
* Change the theme (to make the figure clear).
* <tt>theme_bw()</tt>.
]


.pull-right[
```{r}
library(ggplot2)
gg <- ggplot(mtcars, aes(hp, mpg)) +
        geom_point(aes(color=as.factor(cyl)), size=5) + labs(x = "Horsepower",y= "Miles Per Gallon",color= "# of Cylinders")+
        theme_bw()
gg
```
]

---

# Fourth layer: add the regression line 

.pull-left[
* Add a regression line.
* <tt>geom_smooth(method="lm", se=FALSE)</tt>
]


.pull-right[
```{r}
library(ggplot2)
gg <- ggplot(mtcars, aes(hp, mpg)) +
        geom_point(aes(color=as.factor(cyl)), size=5) + geom_smooth(method="lm", se=FALSE) +
        labs(x = "Horsepower",y= "Miles Per Gallon", color= "# of Cylinders") +  theme_bw()
gg
```
]

---




# Fourth layer: lines by line 


.pull-left[
* <tt>ggplot(data,aes(variables))</tt>.
* <tt>geom_point(size)</tt>.
* <tt>aes(color the data points)</tt>.
* <tt>geom_smooth(add regression line)</tt>.
* <tt>labs(add labels)</tt>.
]


.pull-right[
```{r}
library(ggplot2)
gg <- ggplot(mtcars, aes(hp, mpg)) +
        geom_point(aes(color=as.factor(cyl)), size=5) + geom_smooth(method="lm", se=FALSE) +
        labs(x = "Horsepower",y= "Miles Per Gallon", color= "# of Cylinders") + theme_bw()
gg
```
]

---




# Fourth layer: add C.Is 

.pull-left[
* Add C.Is to the regression line.
* <tt>geom_smooth(method="lm", se=TRUE)</tt>
]


.pull-right[
```{r}
library(ggplot2)
gg <- ggplot(mtcars, aes(hp, mpg)) + geom_point(aes(color=as.factor(cyl)), size=5) +
        geom_smooth(method="lm", se=TRUE) +
        labs(x = "Horsepower",y= "Miles Per Gallon",color= "# of Cylinders") +theme_bw()
gg
```
]

---


# Fifth layer: add a smoother 

.pull-left[
* Add C.I to the regression line.
* <tt>geom_smooth(method="loess"", se=TRUE)</tt>
]


.pull-right[
```{r}
library(ggplot2)
gg <- ggplot(mtcars, aes(hp, mpg)) +geom_point(aes(color=as.factor(cyl)), size=5) +
        geom_smooth(method="lm", se=TRUE) + geom_smooth(method="loess",colour = "blue", size = 1.5)+
        labs(x = "Horsepower",y= "Miles Per Gallon",color= "# of Cylinders") + theme_bw()
gg
```
]

---

# Visualize patterns  across cylinders  

.pull-left[
* How the dependence between mpg and hp with Cylinder numbers.
]


.pull-right[

```{r}
qplot(hp,mpg,data = mtcars, colour = factor(cyl))+
geom_smooth(method = "lm",se = F)
```
]


---



class: center, middle

# Part 3: Location

`Focus: the center of the distribution` <br>

`Graphical and numerical summaries` <br>

---


# YouTube tutorial: calculating Mean, Median, Range, Minimum and Maximum using R studio 

For a short online YouTube tutorial, by BIO-RESEARCH, about the mean, median etc, using R see [YTVD4](https://www.youtube.com/watch?v=nkSl9hwh-dQ).

---

# YouTube tutorial: dotplot

* For a short online YouTube tutorial, by ramstatvid, about dotplot using the \texttt{lattice} package see [YTVD5]( https://www.youtube.com/watch?v=-FGU6PMaSRY).

* For a short online YouTube tutorial, by ramstatvid, about dotplot using the \texttt{gg2plot} package see [YTVD6]( https://www.youtube.com/watch?v=y1CsT-jAWZQ)

---


# Location

.pull-left[
* Location is the center of the distribution. 
* The figure presents distributions with different locations.
* Two normal densities with mean equal to 0 (the black line) and mean equal to 2 (the red line).
]

.pull-right[
```{r}
x<-seq(from=-5,to=5,length=1000)
dx1<-dnorm(x,0,1)
dx2<-dnorm(x,2,1)
plot(x,dx1,type="l")
lines(x,dx2,col=2)
```
]

---

# Location

.pull-left[
* Histograms for two $random$ $samples$ drawn from normal distribution with the same variance but different  mean. 
* Both histograms show that the data are symmetric around the sample mean but the histogram of $x_{2}$ is located to the right relative to the histogram of $x_{1}$. 
]

.pull-right[
```{r}
x1 <- rnorm(10000, 0, 1)
x2 <- rnorm(10000, 2, 1)
par(mfrow = c(2, 1))
hist(x1, col = 0, nclass = 50, xlim = c(-4, 6))
hist(x2, col = 0, nclass = 50, xlim = c(-4, 6))
```
]

---


# Graphical displays for location
.pull-left[
* How can we visualize the distribution ?
* How can we visualize a shift in location ?
* Numerical summaries for location (look at the book): mean. median, trimmed mean...
]

.pull-right[
```{r,echo=FALSE}
x<-seq(from=-5,to=5,length=1000)
dx1<-dnorm(x,0,1)
dx2<-dnorm(x,2,1)
plot(x,dx1,type="l")
lines(x,dx2,col=2)
```
]

---

#  The singer dataset
* The singer datset (the R object <tt>singer</tt>) is a data frame giving the heights of singers in the New York Choral Society. 
* The variables are named <tt>height</tt> (inches) and <tt>voice.part</tt> which is the voice group of the singer

   * Alto.
   * Sporano.
   * Tenor.
   * Bass.
   
* Each voice group is subdivied into two groups, high voice and low voice (for example Bass1 and Bass2 and the lower and higher Bass voices, respectivly). 
* Main focus: the distribution of the singers' <tt>height</tt>.

---

#  The singer dataset

* The data:

```{r}
head(singer)
```

---

# Stripplot (<tt>lattice</tt>)

.pull-left[
* Stripplot:  plots the data of each voice group in a different strip. 
* <tt>voice.part~height</tt>.
* A clear main pattern in the data: 
  * It is easy to distinguish between women (Sopranos and Altos) and men (Tenors and Basses). 
  * Women are clearly shorter than men.
]

.pull-right[
```{r}
dotplot(singer$voice.part~singer$height, 
		aspect=1,
		xlab="Mean Height (inches)")
```
]

---

# Stripplot (<tt>ggplot2</tt>): first layer

.pull-left[
* An equivalent stripplot can be produced using the <tt>ggplot()</tt> package. 
* The function <tt>geom_jitter</tt>: observatios with the same values will be ploted side by side and will not overlap so sample size would be seen as well in the plot.  
]

.pull-right[
```{r}
ggplot(singer, 
       aes(voice.part,height)) + 
       geom_jitter(position = position_jitter(width = .05))
```
]

---

# Mean by voice group

* In order to get a better insight of other patterns, we can summarize the distribution of each group with the sample mean. 
* In R this can be done using the function <tt>tapply()</tt>. 
* To calculate mean of height by the voice group in the singer dataset we use

```{r}
attach(singer)
tapply(singer$height,singer$voice.part,mean)
```

---

# Mean by voice group

* The group means point on the pattern that was already detected: on average men are taller than women. 
* In addition, we can see that within each gender group, singers with lower voice are taller than singers with higher voice. 
* For example, the average of the two bass groups (71.38 and 70.71 for Bass 1 and Bass 2 respectively) are higher than the average in the tenor groups (69.90 and 68.90 for Tenor 1 and Tenor 2 respectively). 

---

# Mean by voice group

* Among women, the sopranos are shorter, on average, than the altos. 
* Within each voice group (all except the sopranos), the singers with lower voices (the second voice group Bass 2, Tenor 2 and Alto 2) are taller than the singers with the higher voices (the first group Bass 1, Tenor 1 and Alto 1). 
* For example the mean of the Bass 2 group (71.38) is higher than the mean of the Bass 1 group (70.71)

---


# Stripplot (<tt>ggplot2</tt>): second layer

.pull-left[
*  The Figure shows the same information as in slide 34 with the addtion of the mean for each voice group.
* <tt>stat_summary()</tt>.
* <tt>fun.y = "mean"</tt>.
]


.pull-right[
```{r}
ggplot(singer, aes(voice.part,height)) +
geom_point() +
stat_summary(geom = "point", fun.y = "mean", colour = "red", size = 4)
```
]

---

# Boxplot (<tt>lattice</tt>)

.pull-left[
* Graphical display of the location of each distribution is the box plot.
* The location of each group is summarized by the median (the dot inside the box). 
* Other aspects of this plot will be discussed in later chapters). 
* Note that the function <tt>bwplot()</tt> is a part of the <tt>lattice</tt> package.
]

.pull-right[
```{r}
bwplot(as.factor(singer$voice.part)~ singer$height,
		data=singer,
		aspect=1,
		xlab="Height (inches)")
```		
]

---

# Multiway histogram (<tt>lattice</tt>)

.pull-left[

* The multiway histogram presents the distributions of heights across the voice groups. 
* Note how the distribution of height is shiffted from left to right across the voice levels.
]

.pull-right[
```{r}
histogram(~ singer$height | singer$voice.part,
		data=singer, layout = c(2, 4), 
		  aspect = 0.5,xlab = "height")
```
]

---

# Multiway histogram (<tt>ggplot2</tt>)

.pull-left[
* An equivalent multiway histogram can be produce with the <tt>ggplot2</tt> package.
* Addtional two layers in the basic plot:
  * The first layer spacifys the plot type <tt>histogram()</tt>.
  * The second layer indicates the factor for the plot partions.
  * The function <tt>facet_wrap(factor)</tt>. 
]


.pull-right[
```{r}
ggplot(singer, aes(height,fill = voice.part)) +
geom_histogram() +
facet_wrap(~voice.part,ncol = 2)
```
]

---


class: center, middle

# Part 4: Spread

`Focus: Variability around the center of the distribution` <br>

---


# Spread

.pull-left[
* Two normal densities that have different variability (or spread). 
* The density with the black line has variance 1 and density with the green line has variance 2.  
]

.pull-right[
```{r}
dx3<-dnorm(x,0,2)
plot(x,dx1,type="l")
lines(x,dx3,col=3)
```
]

---

# Spread

.pull-left[
* Two samples  were drawn from normal distribution with mean equal to 0 but with different variance. 
* The two distributions have the same shape, both histograms are symmetric around 0 as expected. 
* The spread in the histogram of $x_{2}$ is much higher than the spread in the histogram of $x_{1}$.
]

.pull-right[
```{r}
x1 <- rnorm(10000, 0, 0.75)
x2 <- rnorm(10000, 0, 2)
par(mfrow = c(2, 1))
hist(x1, col = 0, nclass = 25, xlim = c(-4, 6), ylim = c(0, 1000))
hist(x2, col = 0, nclass = 50, xlim = c(-4, 6), ylim = c(0, 1000))
```
]

---

# Main concepts
 * Up till now we summarized the distribution of the data with location estimators.
 * In this chapter we focus on the spread. 
 * We want to measure how close the data are to each other and how concentrate the data around the center of the distribution.
 * Numerical summaries for spread (see in the book):
 
    * The sample variance.
    * The fourth-spared.
    * The MAD as measures
  
 * Graphical displays:
 
   * boxplot.
   * violin plot.
   
---


# Boxplot: A graphical display for spread and location

.pull-left[
* Boxplot is a graphical display which shows the location, the spread and the shape of the distribution. 
* The location is summarized by the median, the spread is summarized by the fourth-spread which is simply the length of the box in the boxplot.
* Inside the box: $50\%$ of the data.
]

.pull-right[
```{r, echo=FALSE}
x1<-rnorm(1000,0,1)
par(mfrow=c(2,2))
hist(x1,main="random sample from N(0,1)",xlim=c(-10,10))
boxplot(x1,ylim=c(-10,10))
x2<-rnorm(1000,0,3)
hist(x2,main="random sample from N(0,9)",xlim=c(-10,10))
boxplot(x2,ylim=c(-10,10))
```
]

---

# Boxplot: A graphical display for spread and location

* The upper and lower adjacent values in the boxplot are given by

$$\mbox{Upper adjacent value} = \textit{Min}  \left \{ max(X),Q_{3}+1.5(Q_{3}-Q_{1}) \right \}$$

$$\mbox{Lower adjacent value} = \textit{max}  \left \{ min(X),Q_{1}-1.5(Q_{3}-Q_{1}) \right \}$$

* The upper and lower adjacent values are used to identify extrime values. 
* Observations higher than the upper adjacent value or smaller than the lower adjacent value are considered to be outliers. 

---

# Example: the airqiality data

.pull-left[
* Daily air quality measurements in New York, May to September 1973.
* The histogram and boxplot for the airquality data: Ozone level.
* A skewed distribution with few outliers at the upper tail (histogram). 
* In the boxplot these outliers can be identified above the upper adjacent value.
]

.pull-right[
```{r}
par(mfrow=c(1,2))
airquality1<-na.omit(airquality)
hist(airquality1$Ozone)
boxplot(airquality1$Ozone)
```
]

---


# YouTube tutorial: Boxplot in R

* For a short online YouTube tutorials:

  * by Data Science Tutorials, about boxplot using the <tt>ggplot2</tt> package see
    [YTVD8](https://www.youtube.com/watch?v=48UvtcXTb9U).
    
  * by LawrenceStats, about boxplot  using the <tt>ggplot2</tt> package see 
    [YTVD9](https://www.youtube.com/watch?v=48UvtcXTb9U)

---

# Web tutorial: Advanced boxplots in R

* Example for advanced  boxplots in R using the <tt>ggplot2</tt>  package and code to produce the plots can be found in the R Graph Gallery website here [WAVD2]( https://www.r-graph-gallery.com/boxplot).


---

# Boxplot for the singers data

.pull-left[
* A boxplot for  the singers' height by voice group that was produced using the function <tt>geom_boxplot()</tt>.
]

.pull-right[
```{r}
ggplot(singer, aes(voice.part,height)) + geom_boxplot()
```
]

---

# Boxplot for the singers data


.pull-left[
* The same boxplot in which colors (by group) are added to the boxplot  using the argument <tt>fill=voice.part</tt>. 
* Note that the object <tt>voice.part</tt> is a factor.
]

.pull-right[
```{r}
ggplot(singer, aes(voice.part,height,fill=voice.part)) + geom_boxplot()
```
]

---


# Boxplot for the singers data

.pull-left[
* The data are added to the boxplot using the argument <tt>geom = c("boxplot", "jitter")</tt>.
]

.pull-right[
```{r}
qplot(voice.part, height, data = singer, geom = c("boxplot", "jitter"))
```
]

---

# Violin plot for the singers data

.pull-left[
* When the argument <tt>geom_violin()</tt> is used instead of <tt>geom_boxplot()</tt> the boxplot  become a violin plot.
]

.pull-right[
```{r}
ggplot(singer, aes(voice.part,height)) + geom_violin()
```
]

---




class: center, middle

# Part 5: Shape

`Focus: How does the distribution look like ?` <br>


---



# Shape

.pull-left[
* Two beta densities having diffent shapes.
* Black line: $Beta(2,2)$, red line: $Beta(2,4)$.
]

.pull-right[
```{r figchp12i,echo=FALSE,fig.cap="Shape"}
x<-seq(from=0,to=1,length=1000)
dx1 <- dbeta(x, 2, 2)
dx2 <- dbeta(x, 2, 4)
plot(x,dx1,type="l",ylim=c(0,2.5))
lines(x,dx2,col=2)
```
]

---

# Shape

.pull-left[
*  4 samples (each with 10000 observations) that were drawn from different distributions.
* $x_{1}$ and $x_{2}$: samples were drawn from symmetric distributions. 
* The distributions of $x_{3}$ is skewed to the left and the distribution of $x_{4}$ to the right.
]


.pull-right[
```{r, echo=FALSE}
x1 <- rbeta(10000, 1, 1)
x2 <- rbeta(10000, 2, 2)
x3 <- rbeta(10000, 8, 3)
x4 <- rbeta(10000, 3, 8)
par(mfrow = c(2, 2))
hist(x1, col = 0, nclass = 50)
hist(x2, col = 0, nclass = 50)
hist(x3, col = 0, nclass = 50)
hist(x4, col = 0, nclass = 50)
```
]

---


# Density and density estimate

* So far we used histogram to visualize the shape of the distribution of the observations in the sample. 
* In this chapter we discuss density estimates as a method to estimate and visualized the distribution in the population.  

---

# Density and density estimate

.pull-left[

* A density function of $N(0,1)$ that represents the distribution of a random variable in the population. 
* Suppose that we draw a random sample of size $n$ from the population. 
* The histogram can be used to visualize the shape of the distribution. 
* It is an estimate for the density in the population. 
]

.pull-right[

```{r, echo=FALSE}
x<-seq(from=-3,to=3,length=1000)
dx1<-dnorm(x,0,1)
par(mfrow=c(1,2))
plot(x,dx1,type="l")
title("a")
x1 <- rnorm(1000, 0, 1)
hist(x1, col = 0, nclass = 50, xlim = c(-3.5,3.5),main=" ")
title("b")
```
]


---

# Density and density estimate

.pull-left[
* A second approach to estimate the distribution of the population is to use a smooth version of the histogram
, i.e., a density estimate. 
* The density estimate for our example is shown (in red) panel c and d.
]

.pull-right[
```{r ,echo=FALSE}
x<-seq(from=-3,to=3,length=1000)
dx1<-dnorm(x,0,1)
par(mfrow=c(1,2))
#plot(x,dx1,type="l")
#title("a")
#x1 <- rnorm(1000, 0, 1)
#hist(x1, col = 0, nclass = 50, xlim = c(-3.5,3.5),main=" ")
#title("b")
dx2<-density(x1)
hist(x1, col = 0, nclass = 50, xlim = c(-3.5,3.5),main=" ",probability =T)
lines(dx2$x,dx2$y,col=2,lwd=2)
title("c")
plot(x,dx1,type="l",ylim=c(0,0.55))
lines(dx2$x,dx2$y,col=2,lwd=2)
title("d")
```
]
---


# YouTube tutorial: Creating density plots and enhancing it with the <tt>ggplot2</tt> package 

* A short online YouTube tutorial by LawrenceStats, about density plot using the <tt>ggplot2</tt> package see [YTVD10](https://www.youtube.com/watch?v=kvg0y90GX0Y).

---

# Web tutorial: the ridgeline chart

* A Web tutorial about the ridgeline chart using the <tt>ggplot2</tt> and <tt>ggridges</tt> package is given in the the R Graph Gallery website [WAVD4](https://www.r-graph-gallery.com/ridgeline-plot.html).

---


#  The old faithful data

* The data gives information about waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA.
* Two variables:
  
   * <tt>eruptions</tt>:	numeric,	Eruption time in mins.
   * <tt>waiting</tt>:	numeric,	Waiting time to next eruption (in mins)

---


#  The old faithful data


.pull-left[

* Histogram of eruptions time.
* A bi-model.
* <tt>hist()</tt>: basic graphical function in R.
]


.pull-right[

```{r}
hist(faithful$eruptions,nclass=20,col=2)
```
]

---

#  The old faithful data


.pull-left[

* A failure of the boxplot to capture this feature (the bi-model) of the data .
]


.pull-right[

```{r}
qplot(rep(1,length(eruptions)),eruptions, data=faithful, geom = c("boxplot"))
```
]

---

#  The old faithful data

.pull-left[
* We add the data to the boxplot, using the option <tt>geom = c("boxplot", "jitter")</tt>.
* We identify the two parts of the  eruptions time distribution.
]


.pull-right[
```{r}
qplot(rep(1,length(eruptions)),eruptions, data=faithful, geom = c("boxplot", "jitter"))
```
]

---


#  The old faithful data

.pull-left[
* The histogram is able to capture the shape of the distribution. 
]


.pull-right[

```{r}
qplot(eruptions, data=faithful, geom="histogram")
```
]

---

#  The old faithful data

.pull-left[
* A density estimate was produced using the opion <tt>geom="density")</tt> provides a smooth estimate of the distribuon.
]


.pull-right[

```{r}
qplot(eruptions, data=faithful, geom="density")
```
]

---


#  The old faithful data

.pull-left[

* We can plot both histogram and density in the same plot.
]



.pull-right[
```{r}
hist(faithful$eruptions,nclass=15,probability = TRUE)
dx<-density(faithful$eruptions)
lines(dx$x,dx$y,lwd=2,col=2)
```
]

---

#  The singer data

.pull-left[
* We use density plots to visualize the shit of the distribution of the singers' height across the voice part groups.
* The histograms ravel the shifts within and between the voice groups.
]


.pull-right[
```{r}
ggplot(singer, aes(height,fill = voice.part)) +
geom_histogram() +
facet_wrap(~voice.part,ncol = 2)
```
]

---

#  The singer data

.pull-left[

* In the density plots:
  * The difference between the sopranos and altos (women singers, the densities in the left) and the tenors and basses (men singers, densities in the right) is clearly seen.
  * The difference within each group is more difficult to detect. 
]

.pull-right[
```{r}
qplot(height, data=singer, geom="density", xlim = c(50,80),
fill = voice.part, alpha = I(0.2))
```
]

---

#  The singer data

.pull-left[
* A ridgeline charts visualizes the difference between the groups and within each voice group. 
* Note that the R package <tt>ggridges</tt> should be instaled to produce the plot.
]

.pull-right[
```{r}
library(ggridges)
ggplot(singer, aes(x=height,y=voice.part,fill = voice.part)) +
  geom_density_ridges() +
  theme_ridges() + 
  theme(legend.position = "none")
```
]

---

class: center, middle

# Part 6: Shape (normal probability plot)

`Focus: Normal distribution?` <br>

---


#  Normal probability plot

* A normal probability plot is a plot in which the quantile of the samples are plotted versus the corresponding quantiles of a standard normal distribution $N(0,1)$. 
* In this chapter we discuss the normal probability plot as a graphical tools to vizualise the shape of a distribution. 
* Using histograms and boxplots we are able to investigate the shape of the distribution focusing on the follwing issues:

  * How nearly symmetric the distribution of the data is.
  * Whether the distribution of the data is single-peaked, or whether it is multi-peaked.
  * Whether it is skewed.
  * How far we  from a Normal distribution ?
  
---


# Quantile of $N(\mu,\sigma^{2})$: Definition and a simple example

* A qq normal plot is a graphical disply to investigate how now nearly is the sample to a normal distribution. Let $q_{\mu,\sigma}(f)$ be a quantile of $N(\mu,\sigma^{2})$, it can be expressed as

$$q_{\mu,\sigma}(f)=\mu+\sigma q_{0,1}(f).$$

* For example, the $2.5 \% $ quantile of the standard normal distribution is -1.96. 

```{r}
qnorm(0.025,0,1)
```

* For $N(2,5^{2})$ we have 

$$q_{2,5}(2.5 \% )=2+ 5 \times - 1.96=-7.8$$


```{r}
qnorm(0.025,2,5)
```

---

# YouTube tutorial: QQ-plots in RStudio

For a short online YouTube tutorials by UTSSC, about normal probability plot using R studio see [YTVD12](https://www.youtube.com/watch?v=louihSVsMC0).


---



# A sample from N(0,1)

.pull-left[
* We use the R function <tt>qqnorm()</tt> to produce the normal probabilty plot.
* Data were sampled from $N(0,1)$.
* we expect that all points in the normal probability plot will lay on the on the $45^o$ line. 
]


.pull-right[
```{r}
x <- rnorm(1000, 0, 1)
par(mfrow = c(2, 2))
hist(x, nclass = 25, col = 0)
boxplot(x, boxcol = 2, medcol = 1)
qqnorm(x)
abline(0, 1)
```
]

---




# A Sample from N(2,1)

.pull-left[
* A sample  from $N(2,1)$, i.e., it represents a shift model with the same variabiliy comapre to $N(0,1)$.
* In this case we expect that all points in the normal probability plot will lay above and parallel to the $45^{o}$ lines. 
]

.pull-right[
```{r}
par(mfrow = c(2, 2))
x <- rnorm(1000, 2, 1)
hist(x, nclass = 25, col = 0)
boxplot(x, boxcol = 2, medcol = 1)
qqnorm(x)
abline(0, 1)
```
]

---


# A Sample from N(0,2)

.pull-left[
* The sample was drawn from $N(0,2^2)$ which implies that the mean is the same as $N(0,1)$ but the variability is higher. 
* We expect the points in the normal probability plot to form a straight line with higher slope than 1.
]

.pull-right[
```{r}
par(mfrow = c(2, 2))
x <- rnorm(1000, 0, 2)
hist(x, nclass = 25, col = 0)
boxplot(x, boxcol = 2, medcol = 1)
qqnorm(x)
abline(0, 1)
```
]


---


# A Sample from $t_{(3)}$

.pull-left[

* A $t_{(3)}$ distribution (red line) has the same mean as $N(0,1)$ but longer tails. 
* Note that the two distribution and centered around zero. 
]


.pull-right[

```{r, echo=FALSE}
par(mfrow = c(1, 1))
```


```{r,echo=FALSE}
qx <- seq(from = -7, to = 7, length = 1000)
xn <- dnorm(qx, mean = 0, sd = 1)
xt <- dt(qx, 3)
plot(qx, xn, xlim = c(-7, 7), type = "l")
lines(qx, xt, col=2)
```
]

---

# A Sample from $t_{(3)}$

.pull-left[
* For the normal probability plot we expect that the points will lay on the $45^o$ line in the center but with more extreme values .
]


.pull-right[
```{r}
par(mfrow = c(2, 2))
x <- rt(1000, 3)
hist(x, nclass = 25, col = 0)
boxplot(x, boxcol = 2, medcol = 1)
qqnorm(x)
abline(0, 1)
```
]

---




# A Sample from $t_{(3)}$


```{r,echo=FALSE}
par(mfrow = c(2, 2))
x <- rt(1000, 3)
hist(x, nclass = 25, col = 0)
boxplot(x, boxcol = 2, medcol = 1)
qqnorm(x)
abline(0, 1)
```

---


# A Sample from $U(-3,3)$

.pull-left[
* The data of this example are uniformly distributed across the minimum and maximum values. 
* We expect the points in the normal probability plot to cross the $45^o$ lines and to lay relatively far from the line.
]


.pull-right[
```{r}
x <- runif(1000, -3, 3)
par(mfrow = c(2, 2))
hist(x, nclass = 25, col = 0)
boxplot(x, boxcol = 2, medcol = 1)
qqnorm(x)
abline(0, 1)
```
]

---


# A Sample from $U(-3,3)$

```{r,echo=FALSE}
x <- runif(1000, -3, 3)
par(mfrow = c(2, 2))
hist(x, nclass = 25, col = 0)
boxplot(x, boxcol = 2, medcol = 1)
qqnorm(x)
abline(0, 1)
```

---





#  The cars data

.pull-left[

* Normal probability plot for for mpg.
* $N(0,1)$ ?
* $N(\mu,\sigma^{2})$ ?
]



.pull-right[
```{r}
qqnorm(mtcars$mpg)
```
]

---



#  The cars data

.pull-left[
* Normal probability plot for for the z-score  of mpg.
* $N(0,1)$ ?
]



.pull-right[
```{r}
m.mpg<-mean(mtcars$mpg)
sd.mpg<-sqrt(var(mtcars$mpg))
z<-(mtcars$mpg-m.mpg)/sd.mpg
qqnorm(z)
abline(0,1,col=2)
```
]

---


# The signer dataset

.pull-left[
* Normal probability plot for the height.
]

.pull-right[
```{r}
qqmath(~ height,
		distribution = qnorm,
		data=singer,
	       layout=c(1,1), 
             prepanel = prepanel.qqmathline,
            panel = function(x, ...) {
            panel.grid()
            panel.qqmathline(x, ...)
            panel.qqmath(x, ...)
            },
		aspect=1,
		xlab = "f-value",
		ylab="height")
```
]

---




# The signer dataset


```{r, echo=FALSE}
qqmath(~ height,
		distribution = qnorm,
		data=singer,
	       layout=c(1,1), 
             prepanel = prepanel.qqmathline,
            panel = function(x, ...) {
            panel.grid()
            panel.qqmathline(x, ...)
            panel.qqmath(x, ...)
            },
		aspect=1,
		xlab = "f-value",
		ylab="height")
```

---



# The signer dataset

.pull-left[
* Normal probability plot for the height (by voice group).
]

.pull-right[
```{r}
qqmath(~ height | voice.part,
		distribution = qnorm,
		data=singer,
	       layout=c(4,2), 
             prepanel = prepanel.qqmathline,
            panel = function(x, ...) {
            panel.grid()
            panel.qqmathline(x, ...)
            panel.qqmath(x, ...)
            },
		aspect=1,
		xlab = "f-value",
		ylab="height")
```
]

---


# The signer dataset

```{r,echo=FALSE}
qqmath(~ height | voice.part,
		distribution = qnorm,
		data=singer,
	       layout=c(4,2), 
             prepanel = prepanel.qqmathline,
            panel = function(x, ...) {
            panel.grid()
            panel.qqmathline(x, ...)
            panel.qqmath(x, ...)
            },
		aspect=1,
		xlab = "f-value",
		ylab="height")
```

---


